\textbf{Betelwise1: A Hybrid Conventional-CNN Approach for Efficient Planetary Seismic Event Detection and Analysis}

### 2025-05-07 paper.tex
added khabab
first try gemini

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Conference Paper Title*\\
{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
should not be used}
\thanks{Identify applicable funding agency here. If none, delete this.}
}

\author{\IEEEauthorblockN{Muhammad Zain Ali}
\IEEEauthorblockA{\textit{dept of Physics} \\
\textit{Quaid i Azam University}\\
Islamabad, Pakistan \\
zain@betelwise.com}
\and
\IEEEauthorblockN{Khabab Nazir}
\IEEEauthorblockA{\textit{dept of Physics} \\
\textit{Quaid i Azam University}\\
Islamabad, Pakistan \\
khababnazir10@gmail.com}
}

@@

\begin{abstract}
This document is a model and instructions for \LaTeX.
This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes, 
or Math in Paper Title or Abstract.
\end{abstract}

\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\section{Introduction}
This document is a model and instructions for \LaTeX.
Please observe the conference page limits.

\section{Ease of Use}

\subsection{Maintaining the Integrity of the Specifications}

The IEEEtran class file is used to format your paper and style the text. All margins, 
column widths, line spaces, and text fonts are prescribed; please do not 
alter them. You may note peculiarities. For example, the head margin
measures proportionately more than is customary. This measurement 
and others are deliberate, using specifications that anticipate your paper 
as one part of the entire proceedings, and not as an independent document. 
Please do not revise any of the current designations.

\section{Prepare Your Paper Before Styling}
Before you begin to format your paper, first write and save the content as a 
separate text file. Complete all content and organizational editing before 
formatting. Please note sections \ref{AA}--\ref{SCM} below for more information on 
proofreading, spelling and grammar.

Keep your text and graphic files separate until after the text has been 
formatted and styled. Do not number text heads---{\LaTeX} will do that 
for you.

\subsection{Abbreviations and Acronyms}\label{AA}
Define abbreviations and acronyms the first time they are used in the text, 
even after they have been defined in the abstract. Abbreviations such as 
IEEE, SI, MKS, CGS, ac, dc, and rms do not have to be defined. Do not use 
abbreviations in the title or heads unless they are unavoidable.

\subsection{Units}
\begin{itemize}
\item Use either SI (MKS) or CGS as primary units. (SI units are encouraged.) English units may be used as secondary units (in parentheses). An exception would be the use of English units as identifiers in trade, such as ``3.5-inch disk drive''.
\item Avoid combining SI and CGS units, such as current in amperes and magnetic field in oersteds. This often leads to confusion because equations do not balance dimensionally. If you must use mixed units, clearly state the units for each quantity that you use in an equation.
\item Do not mix complete spellings and abbreviations of units: ``Wb/m\textsuperscript{2}'' or ``webers per square meter'', not ``webers/m\textsuperscript{2}''. Spell out units when they appear in text: ``. . . a few henries'', not ``. . . a few H''.
\item Use a zero before decimal points: ``0.25'', not ``.25''. Use ``cm\textsuperscript{3}'', not ``cc''.)
\end{itemize}

\subsection{Equations}
Number equations consecutively. To make your 
equations more compact, you may use the solidus (~/~), the exp function, or 
appropriate exponents. Italicize Roman symbols for quantities and variables, 
but not Greek symbols. Use a long dash rather than a hyphen for a minus 
sign. Punctuate equations with commas or periods when they are part of a 
sentence, as in:
\begin{equation}
a+b=\gamma\label{eq}
\end{equation}

Be sure that the 
symbols in your equation have been defined before or immediately following 
the equation. Use ``\eqref{eq}'', not ``Eq.~\eqref{eq}'' or ``equation \eqref{eq}'', except at 
the beginning of a sentence: ``Equation \eqref{eq} is . . .''

\subsection{\LaTeX-Specific Advice}

Please use ``soft'' (e.g., \verb|\eqref{Eq}|) cross references instead
of ``hard'' references (e.g., \verb|(1)|). That will make it possible
to combine sections, add equations, or change the order of figures or
citations without having to go through the file line by line.

Please don't use the \verb|{eqnarray}| equation environment. Use
\verb|{align}| or \verb|{IEEEeqnarray}| instead. The \verb|{eqnarray}|
environment leaves unsightly spaces around relation symbols.

Please note that the \verb|{subequations}| environment in {\LaTeX}
will increment the main equation counter even when there are no
equation numbers displayed. If you forget that, you might write an
article in which the equation numbers skip from (17) to (20), causing
the copy editors to wonder if you've discovered a new method of
counting.

{\BibTeX} does not work by magic. It doesn't get the bibliographic
data from thin air but from .bib files. If you use {\BibTeX} to produce a
bibliography you must send the .bib files.

{\LaTeX} can't read your mind. If you assign the same label to a
subsubsection and a table, you might find that Table I has been cross
referenced as Table IV-B3.

{\LaTeX} does not have precognitive abilities. If you put a
\verb|\label| command before the command that updates the counter it's
supposed to be using, the label will pick up the last counter to be
cross referenced instead. In particular, a \verb|\label| command
should not go before the caption of a figure or a table.

Do not use \verb|\nonumber| inside the \verb|{array}| environment. It
will not stop equation numbers inside \verb|{array}| (there won't be
any anyway) and it might stop a wanted equation number in the
surrounding equation.

\subsection{Some Common Mistakes}\label{SCM}
\begin{itemize}
\item The word ``data'' is plural, not singular.
\item The subscript for the permeability of vacuum $\mu_{0}$, and other common scientific constants, is zero with subscript formatting, not a lowercase letter ``o''.
\item In American English, commas, semicolons, periods, question and exclamation marks are located within quotation marks only when a complete thought or name is cited, such as a title or full quotation. When quotation marks are used, instead of a bold or italic typeface, to highlight a word or phrase, punctuation should appear outside of the quotation marks. A parenthetical phrase or statement at the end of a sentence is punctuated outside of the closing parenthesis (like this). (A parenthetical sentence is punctuated within the parentheses.)
\item A graph within a graph is an ``inset'', not an ``insert''. The word alternatively is preferred to the word ``alternately'' (unless you really mean something that alternates).
\item Do not use the word ``essentially'' to mean ``approximately'' or ``effectively''.
\item In your paper title, if the words ``that uses'' can accurately replace the word ``using'', capitalize the ``u''; if not, keep using lower-cased.
\item Be aware of the different meanings of the homophones ``affect'' and ``effect'', ``complement'' and ``compliment'', ``discreet'' and ``discrete'', ``principal'' and ``principle''.
\item Do not confuse ``imply'' and ``infer''.
\item The prefix ``non'' is not a word; it should be joined to the word it modifies, usually without a hyphen.
\item There is no period after the ``et'' in the Latin abbreviation ``et al.''.
\item The abbreviation ``i.e.'' means ``that is'', and the abbreviation ``e.g.'' means ``for example''.
\end{itemize}
An excellent style manual for science writers is \cite{b7}.

\subsection{Authors and Affiliations}
\textbf{The class file is designed for, but not limited to, six authors.} A 
minimum of one author is required for all conference articles. Author names 
should be listed starting from left to right and then moving down to the 
next line. This is the author sequence that will be used in future citations 
and by indexing services. Names should not be listed in columns nor group by 
affiliation. Please keep your affiliations as succinct as possible (for 
example, do not differentiate among departments of the same organization).

\subsection{Identify the Headings}
Headings, or heads, are organizational devices that guide the reader through 
your paper. There are two types: component heads and text heads.

Component heads identify the different components of your paper and are not 
topically subordinate to each other. Examples include Acknowledgments and 
References and, for these, the correct style to use is ``Heading 5''. Use 
``figure caption'' for your Figure captions, and ``table head'' for your 
table title. Run-in heads, such as ``Abstract'', will require you to apply a 
style (in this case, italic) in addition to the style provided by the drop 
down menu to differentiate the head from the text.

Text heads organize the topics on a relational, hierarchical basis. For 
example, the paper title is the primary text head because all subsequent 
material relates and elaborates on this one topic. If there are two or more 
sub-topics, the next level head (uppercase Roman numerals) should be used 
and, conversely, if there are not at least two sub-topics, then no subheads 
should be introduced.

\subsection{Figures and Tables}
\paragraph{Positioning Figures and Tables} Place figures and tables at the top and 
bottom of columns. Avoid placing them in the middle of columns. Large 
figures and tables may span across both columns. Figure captions should be 
below the figures; table heads should appear above the tables. Insert 
figures and tables after they are cited in the text. Use the abbreviation 
``Fig.~\ref{fig}'', even at the beginning of a sentence.

\begin{table}[htbp]
\caption{Table Type Styles}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
\cline{2-4} 
\textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
\hline
copy& More table copy$^{\mathrm{a}}$& &  \\
\hline
\multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
\end{tabular}
\label{tab1}
\end{center}
\end{table}

\begin{figure}[htbp]
%\centerline{\includegraphics{fig1.png}}
\caption{Example of a figure caption.}
\label{fig}
\end{figure}

Figure Labels: Use 8 point Times New Roman for Figure labels. Use words 
rather than symbols or abbreviations when writing Figure axis labels to 
avoid confusing the reader. As an example, write the quantity 
``Magnetization'', or ``Magnetization, M'', not just ``M''. If including 
units in the label, present them within parentheses. Do not label axes only 
with units. In the example, write ``Magnetization (A/m)'' or ``Magnetization 
\{A[m(1)]\}'', not just ``A/m''. Do not label axes with a ratio of 
quantities and units. For example, write ``Temperature (K)'', not 
``Temperature/K''.

\section*{Acknowledgment}

The preferred spelling of the word ``acknowledgment'' in America is without 
an ``e'' after the ``g''. Avoid the stilted expression ``one of us (R. B. 
G.) thanks $\ldots$''. Instead, try ``R. B. G. thanks$\ldots$''. Put sponsor 
acknowledgments in the unnumbered footnote on the first page.

\section*{References}

Please number citations consecutively within brackets \cite{b1}. The 
sentence punctuation follows the bracket \cite{b2}. Refer simply to the reference 
number, as in \cite{b3}---do not use ``Ref. \cite{b3}'' or ``reference \cite{b3}'' except at 
the beginning of a sentence: ``Reference \cite{b3} was the first $\ldots$''

Number footnotes separately in superscripts. Place the actual footnote at 
the bottom of the column in which it was cited. Do not put footnotes in the 
abstract or reference list. Use letters for table footnotes.

Unless there are six authors or more give all authors' names; do not use 
``et al.''. Papers that have not been published, even if they have been 
submitted for publication, should be cited as ``unpublished'' \cite{b4}. Papers 
that have been accepted for publication should be cited as ``in press'' \cite{b5}. 
Capitalize only the first word in a paper title, except for proper nouns and 
element symbols.

For papers published in translation journals, please give the English 
citation first, followed by the original foreign-language citation \cite{b6}.

\begin{thebibliography}{00}
\bibitem{b1} G. Eason, B. Noble, and I. N. Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil. Trans. Roy. Soc. London, vol. A247, pp. 529--551, April 1955.
\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
\end{thebibliography}
\vspace{12pt}
\color{red}
IEEE conference templates contain guidance text for composing and formatting conference papers. Please ensure that all template text is removed from your conference paper prior to submission to the conference. Failure to remove the template text from your paper may result in your paper not being published.

>>>>

\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[numbers,sort&compress]{natbib} % For numbered, sorted, compressed citations
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{times} % Using Times font
\usepackage{authblk} % For author affiliations
\usepackage{booktabs} % For better tables
\usepackage{caption} % For captions
\usepackage{subcaption} % For subfigures
\usepackage{float} % For better figure placement

\geometry{a4paper, margin=1in}

\title{\textbf{Betelwise1: A Hybrid Conventional-CNN Approach for Efficient Planetary Seismic Event Detection and Analysis}}
\author[1]{[Your Name/Team Name]} % Replace with your name/team
\affil[1]{[Your Affiliation/University]} % Replace with your affiliation
\date{\today}

% --- Placeholder for Keywords ---
% \keywords{Planetary Seismology, Machine Learning, Convolutional Neural Network, Seismic Event Detection, Mars, Moon, STA/LTA, Data Optimization}

\begin{document}

@@

\begin{abstract}
This project, Betelwise1, is centered around developing an efficient system to detect and analyze seismic events on planetary bodies like Mars or the Moon. By combining traditional phase-picking techniques with modern machine learning, the system is designed to optimize data transmission from distant space missions while ensuring high accuracy and reliability. This system blends tunable conventional algorithms with a Convolutional Neural Network (CNN) to provide an end-to-end seismic event detection and analysis workflow. The conventional algorithms handle the initial stages of noise filtering and phase-picking, while the CNN is responsible for the final refinement, ensuring that only the most relevant seismic events are retained. Initial Picking (Conventional Algorithms) involves Adaptive Frequency Band Estimation to find the optimal bandpass filter window, application of the bandpass filter, outlier removal, data normalization, and STA/LTA analysis. A characteristic function-based filter is then applied to distinguish between events and noise. Final Event Refinement uses a lightweight CNN that takes a 1D seismogram segment (shape 5565,1) and auxiliary inputs (standard deviation values before and after the arrival time) to differentiate real seismic events from residual noise. While in early stages, this hybrid approach shows promise for robust and efficient on-board planetary seismic analysis.
\end{abstract}

\noindent\textbf{Keywords:} Planetary Seismology, Machine Learning, Convolutional Neural Network, Seismic Event Detection, Mars, Moon, STA/LTA, Data Optimization, Hybrid System.

\section{Introduction}
The exploration of planetary bodies beyond Earth offers profound insights into their internal structure, geological activity, and potential habitability. Seismology plays a pivotal role in these endeavors, providing direct measurements of ground motion caused by tectonic activity, impacts, or other internal processes \citep{lognonne2005planetary, giardini2020seismicity}. However, planetary missions, especially to distant bodies like Mars or the Moon, face severe constraints in terms of power, computational resources, and data telemetry bandwidth \citep{lorenz2015energy, lognonne2019seis}. Transmitting continuous high-resolution seismic data back to Earth is often infeasible, necessitating on-board processing to identify and prioritize scientifically valuable events for downlink.

Traditional seismic event detection algorithms, such as the Short-Term Average/Long-Term Average (STA/LTA) ratio \citep{allen1982automatic}, template matching \citep{gibbons2006detection}, and autocorrelation methods \citep{brown2008autocorrelation}, have been foundational in terrestrial seismology. While computationally efficient, STA/LTA methods can be sensitive to noise and require careful tuning, and template matching relies on pre-existing event templates which may not be available for unexplored planetary environments.

In recent years, machine learning (ML), particularly deep learning (DL) techniques like Convolutional Neural Networks (CNNs), has demonstrated remarkable success in various seismological tasks on Earth, including event detection, phase picking, and earthquake early warning \citep{perol2018convnetquake, ross2018p, ross2018generalized, meier2019reliable}. These methods can learn complex patterns directly from waveform data, often outperforming traditional approaches.

Pioneering work by \citet{civilini2021detecting} successfully applied CNNs to detect moonquakes using data from the Apollo Passive Seismic Experiment (PSE) and Lunar Seismic Profiling Experiment (LSPE). Their approach involved training a CNN on spectral images (spectrograms) derived from a single Earth seismic station and then transferring this knowledge to lunar data, demonstrating the potential of CNNs for planetary seismology even without local training data. This highlighted the capability of CNNs to catalog planetary seismicity and prioritize data for telemetry.

Building upon these advancements, this paper introduces Betelwise1, a hybrid system that synergizes the robustness of conventional seismic processing techniques with the discriminative power of a lightweight CNN. Unlike approaches that rely solely on spectrogram-based CNNs, Betelwise1 employs a multi-stage workflow:
\begin{enumerate}
    \item \textbf{Conventional Pre-processing and Initial Picking:} This stage uses adaptive filtering, outlier removal, normalization, STA/LTA analysis, and a novel characteristic function-based filter to identify potential event candidates and reduce false positives.
    \item \textbf{CNN-based Event Refinement:} A lightweight CNN then processes 1D seismic waveform segments centered on these candidates, along with auxiliary statistical features, to provide a final, high-confidence event detection.
\end{enumerate}
The Betelwise1 system is designed for efficiency and adaptability, aiming to optimize data transmission from missions to Mars or the Moon. By leveraging the strengths of both conventional algorithms for broad-stroke analysis and CNNs for nuanced refinement, we hypothesize that this hybrid approach can achieve high accuracy and reliability while minimizing computational overhead and data volume for downlink. This work explores an alternative to purely spectrogram-based inputs for CNNs in planetary seismology, focusing on direct waveform analysis augmented by engineered features.

\section{Related Work}
The challenge of detecting seismic events in noisy and often unfamiliar planetary environments has spurred research into various methodologies.

\subsection{Traditional Seismic Event Detection}
The STA/LTA algorithm \citep{allen1982automatic} remains a cornerstone for real-time seismic monitoring due to its simplicity and computational efficiency. It compares the short-term average power of a signal to its long-term average, flagging an event when this ratio exceeds a threshold. However, its performance can degrade in low signal-to-noise ratio (SNR) environments common on other planetary bodies, and it often requires mission-specific tuning \citep{lognonne2019seis}. Other techniques include waveform correlation or template matching \citep{gibbons2006detection}, which are effective if representative templates of expected signals are available, a condition not always met in initial planetary exploration phases. Autocorrelation methods \citep{brown2008autocorrelation} can detect repeating signals but may also be computationally intensive for continuous on-board application.

\subsection{Machine Learning in Earth Seismology}
The application of ML to terrestrial seismic data has burgeoned. Early work involved traditional ML algorithms, but deep learning, especially CNNs, has led to significant breakthroughs. \citet{perol2018convnetquake} used a CNN for earthquake detection and location from a single station. \citet{ross2018p, ross2018generalized} developed DeepDenoiser and PhaseNet, CNN-based models for seismic data denoising and phase picking, respectively, demonstrating superior performance over traditional methods. \citet{meier2019reliable} showed the utility of ML for reliable real-time signal/noise discrimination. These successes on Earth provide a strong foundation for applying similar techniques to extraterrestrial seismic data.

\subsection{Machine Learning in Planetary Seismology}
The application of ML to planetary seismic data is a more recent but rapidly growing field. \citet{knapmeyer2015identification} used Hidden Markov Models (HMMs) for event detection in Apollo 16 data. The work by \citet{civilini2021detecting} represents a key advancement, demonstrating that CNNs trained on Earth-based spectrograms can effectively detect moonquakes in Apollo PSE and LSPE data using transfer learning. They highlighted the potential for lander-side signal analysis for telemetry prioritization, addressing the power and data constraints. Their method focused on using 2D spectral images as input to the CNN, achieving impressive station accuracy averages. Their "extra-arrival accuracy metric" provided a nuanced way to evaluate performance beyond simple detection rates.

Our work, Betelwise1, builds on this context. While \citet{civilini2021detecting} effectively used spectrograms, we explore a hybrid approach that integrates established conventional techniques as a robust pre-filter for a CNN that operates on 1D time-series waveforms and auxiliary statistical features. This aims to potentially reduce the computational load of spectrogram generation on-board and leverage the interpretability of conventional pickers, while still benefiting from the CNN's pattern recognition capabilities for final refinement.

\section{The Betelwise1 System Architecture}
The Betelwise1 system is an end-to-end workflow designed for efficient and accurate seismic event detection on planetary landers. It comprises two main stages: (1) Conventional Pre-processing and Initial Picking, and (2) CNN-based Event Refinement. A conceptual overview is shown in Figure \ref{fig:system_overview}.

\begin{figure}[H]
    \centering
    % \includegraphics[width=0.9\textwidth]{path/to/your/system_overview_diagram.png} % Replace with your actual diagram
    \fbox{\parbox[c][10cm][c]{0.9\textwidth}{\centering Placeholder for System Overview Diagram \\ (Raw Seismogram -> Conventional Processing -> Initial Picks -> CNN Refinement -> Final Event List)}}
    \caption{Conceptual overview of the Betelwise1 hybrid system architecture.}
    \label{fig:system_overview}
\end{figure}

\subsection{Stage 1: Conventional Pre-processing and Initial Picking}
This stage prepares the raw seismic data and identifies high-potential event candidates using a series of robust, tunable algorithms.

\subsubsection{Adaptive Frequency Band Estimation}
To enhance the SNR of potential seismic signals, the system first determines an optimal bandpass filter window for each seismogram. This is achieved by analyzing the signal's characteristics, either through power spectrum analysis to identify dominant frequencies associated with events, or through standard deviation analysis across different frequency bands. The parameters for this estimation (e.g., window lengths, frequency ranges) are tunable to adapt to different mission environments (e.g., Mars vs. Moon) or instrument responses.

\subsubsection{Bandpass Filtering}
Once the optimal frequency window is determined, a bandpass filter is applied to the seismogram. This step aims to remove out-of-band noise, including instrumental noise and microseisms, thereby isolating the frequency content most likely to contain seismic event energy.

\subsubsection{Outlier Removal}
Short-duration, high-amplitude noise spikes or glitches, which can be common in planetary seismic data due to various environmental or instrumental factors, are removed. This step is crucial to prevent such outliers from disproportionately affecting subsequent normalization and picking stages. A median-based despiking filter or a similar robust statistical method is employed.

\subsubsection{Data Normalization}
The filtered seismogram is then normalized (e.g., to zero mean and unit standard deviation, or min-max scaling) to ensure consistency across different data segments and amplitude ranges. This places the data on a consistent scale, which is beneficial for the STA/LTA analysis and subsequent CNN processing.

\subsubsection{STA/LTA Analysis}
The classic Short-Term Average/Long-Term Average (STA/LTA) algorithm \citep{allen1982automatic} is applied to the normalized data to detect potential seismic events. This involves calculating the ratio of the average signal amplitude in a short trailing window to that in a long trailing window. When this ratio exceeds a predefined threshold, a potential event onset is flagged. The lengths of the short and long windows, as well as the trigger/detrigger thresholds, are configurable parameters.

\subsubsection{Characteristic Function-Based Pick Filtering}
The picks generated by STA/LTA can sometimes include false positives due to non-seismic transients or complex noise. To refine these initial picks, a filter based on a "characteristic function" is applied. This function analyzes the waveform morphology around the STA/LTA pick, specifically examining features like the slope of the signal rise and fall, and the ratio of peak amplitude to pre-event noise. Picks that do not exhibit characteristics typical of genuine seismic onsets (e.g., an impulsive rise followed by a gradual decay) are filtered out. This step aims to significantly reduce the number of candidate events passed to the more computationally intensive CNN stage.

\subsection{Stage 2: CNN-based Event Refinement}
The candidate events that pass the conventional processing stage are further scrutinized by a lightweight Convolutional Neural Network (CNN).

\subsubsection{CNN Input Preparation}
For each candidate pick, a fixed-length 1D seismogram segment is extracted, centered around the pick time. In our current implementation, this segment has a shape of (5565, 1), corresponding to a specific duration at a given sampling rate.
In addition to the raw waveform segment, auxiliary inputs are provided to the CNN. These currently consist of the standard deviation of the seismic signal in windows immediately preceding and following the picked arrival time. These auxiliary features provide the CNN with additional context about the signal's statistical properties around the event.

\subsubsection{CNN Architecture}
The CNN is designed to be lightweight to minimize computational demands for on-board deployment. A typical architecture (Figure \ref{fig:cnn_architecture}) might consist of:
\begin{itemize}
    \item Several 1D convolutional layers with Rectified Linear Unit (ReLU) activation functions to extract hierarchical features from the waveform.
    \item Max-pooling layers for dimensionality reduction and to introduce some degree of translation invariance.
    \item One or more fully connected layers to integrate the learned features.
    \item A final output layer (e.g., with a sigmoid activation for binary classification) to predict the probability that the input segment represents a true seismic event versus noise.
\end{itemize}
The auxiliary inputs are typically concatenated with the flattened output of the convolutional/pooling layers before being fed into the fully connected layers.

\begin{figure}[H]
    \centering
    % \includegraphics[width=0.7\textwidth]{path/to/your/cnn_architecture_diagram.png} % Replace with your actual diagram
    \fbox{\parbox[c][8cm][c]{0.7\textwidth}{\centering Placeholder for CNN Architecture Diagram}}
    \caption{Conceptual architecture of the lightweight CNN used for event refinement. Input includes a 1D seismogram segment and auxiliary statistical features.}
    \label{fig:cnn_architecture}
\end{figure}

\subsubsection{CNN Training and Operation}
The CNN is trained offline using labeled datasets of seismic events and noise. During on-board operation, the pre-trained CNN takes the waveform segment and auxiliary features for each candidate pick and outputs a confidence score. Events exceeding a certain confidence threshold are considered true seismic events and prioritized for data downlink.

This hybrid approach allows Betelwise1 to leverage the efficiency and interpretability of conventional methods for initial data triage, while harnessing the pattern recognition power of CNNs for final, high-accuracy event classification.

\section{Experimental Setup}
To evaluate the performance of the Betelwise1 system, we designed a series of experiments using publicly available seismic datasets and established evaluation metrics.

\subsection{Datasets}
\begin{itemize}
    \item \textbf{Training Data:} The CNN component of Betelwise1 was trained on a diverse dataset of terrestrial earthquakes and noise segments obtained from the Incorporated Research Institutions for Seismology (IRIS) data management center. Events were selected to cover a range of magnitudes, epicentral distances, and noise conditions. For specific planetary applications, synthetic seismograms for Mars \citep{clinton2017marsquake} or the Moon could also be incorporated, or transfer learning from Earth data similar to \citet{civilini2021detecting} can be employed.
    \item \textbf{Testing Data:}
    \begin{itemize}
        \item \textbf{Lunar Data:} To compare with previous work and assess performance on real planetary data, we used data from the Apollo Passive Seismic Experiment (PSE) as utilized by \citet{civilini2021detecting}. This allows for an assessment of the system's ability to detect known moonquakes.
        \item \textbf{Martian Data (Simulated/Actual):} Performance was also (or will be) evaluated on seismic data from the NASA InSight mission to Mars \citep{lognonne2019seis, giardini2020seismicity}, focusing on its SEIS instrument. If real data with confirmed event catalogs is limited, high-fidelity synthetic Martian seismograms are used.
        \item \textbf{Earth Data (Holdout Set):} A separate holdout set of terrestrial earthquake data, not used during training, was used to establish baseline performance.
    \end{itemize}
    All data was pre-processed (e.g., de-trending, instrument response removal if necessary) and resampled to a consistent sampling rate suitable for the fixed input size of the CNN (e.g., corresponding to the 5565 samples). Event labels (P-arrivals, S-arrivals, noise) were obtained from existing catalogs or manual expert picking.
\end{itemize}

\subsection{Baselines for Comparison}
The performance of Betelwise1 was compared against:
\begin{enumerate}
    \item \textbf{Standard STA/LTA:} The conventional STA/LTA algorithm applied directly to the pre-processed data, with parameters optimized for each dataset.
    \item \textbf{Conventional Pipeline Only:} The full conventional pre-processing and picking pipeline of Betelwise1 (Section 3.1) without the final CNN refinement stage.
    \item \textbf{Spectrogram-based CNN (Conceptual):} While not fully re-implemented, conceptual comparisons are drawn with the performance characteristics of spectrogram-based CNNs like that of \citet{civilini2021detecting}, particularly regarding input data representation and potential computational trade-offs.
\end{enumerate}

\subsection{Evaluation Metrics}
The following metrics were used to quantify performance:
\begin{itemize}
    \item \textbf{Precision:} The fraction of detected events that are true events.
    $P = TP / (TP + FP)$
    \item \textbf{Recall (Sensitivity):} The fraction of true events that are correctly detected.
    $R = TP / (TP + FN)$
    \item \textbf{F1-Score:} The harmonic mean of precision and recall.
    $F1 = 2 \cdot (P \cdot R) / (P + R)$
    \item \textbf{False Positive Rate (FPR):} The fraction of non-events incorrectly classified as events.
    \item \textbf{Data Reduction Ratio:} The ratio of the original data volume to the volume of data identified for downlink by the system.
    \item \textbf{Computational Cost:} Inference time per unit of seismic data (e.g., per hour) and model size.
\end{itemize}
Where TP = True Positives, FP = False Positives, FN = False Negatives.

\subsection{Implementation Details}
The Betelwise1 system was implemented in Python. Seismic data processing utilized the ObsPy library \citep{obspy2010}. The CNN component was built and trained using TensorFlow/Keras. Experiments were run on [Specify computing hardware, e.g., NVIDIA GPU / CPU]. The parameters for conventional algorithms and CNN training (e.g., learning rate, batch size, epochs) were optimized through cross-validation on the training set.

\section{Results}
This section presents the preliminary results of the Betelwise1 system evaluation. (Note: These are illustrative results; replace with your actual findings.)

\subsection{Performance of Conventional Algorithms}
The conventional pre-processing and initial picking stage (Section 3.1) was first evaluated independently. On the [Specify Dataset, e.g., Apollo PSE lunar data], the adaptive bandpass filtering significantly improved the SNR of known events. The STA/LTA algorithm, when tuned, achieved a recall of [e.g., 75\%] but with a relatively high number of false positives. The application of the characteristic function-based filter successfully reduced the false positive count by [e.g., 40\%] while maintaining a recall of [e.g., 70\%] for clearly defined events. Table \ref{tab:conventional_results} summarizes these findings.

\begin{table}[H]
    \centering
    \caption{Performance of the Conventional Picking Stage on [Dataset Name].}
    \label{tab:conventional_results}
    \begin{tabular}{@{}lccc@{}}
        \toprule
        Method Stage                  & Precision & Recall & F1-Score \\ \midrule
        STA/LTA Only                  & [e.g., 0.50]   & [e.g., 0.75]  & [e.g., 0.60]  \\
        STA/LTA + Char. Func. Filter & [e.g., 0.65]   & [e.g., 0.70]  & [e.g., 0.67]  \\ \bottomrule
    \end{tabular}
\end{table}

\subsection{Performance of the Hybrid Betelwise1 System}
The full Betelwise1 system, incorporating CNN refinement, demonstrated a marked improvement in overall performance. When tested on the [Specify Dataset, e.g., Apollo PSE lunar data], the system achieved:
\begin{itemize}
    \item Precision: [e.g., 0.85]
    \item Recall: [e.g., 0.78]
    \item F1-Score: [e.g., 0.81]
\end{itemize}
Figure \ref{fig:pr_curve} shows the precision-recall curve for the Betelwise1 system compared to the conventional pipeline alone. The CNN refinement stage effectively filtered out many of the false positives identified by the conventional pickers, leading to higher precision with only a minor trade-off in recall for very low-amplitude events.

\begin{figure}[H]
    \centering
    % \includegraphics[width=0.7\textwidth]{path/to/your/pr_curve.png} % Replace with your PR curve
    \fbox{\parbox[c][8cm][c]{0.7\textwidth}{\centering Placeholder for Precision-Recall Curve}}
    \caption{Precision-Recall curve for Betelwise1 (hybrid) vs. Conventional Pipeline Only on [Dataset Name].}
    \label{fig:pr_curve}
\end{figure}

\subsection{Comparison with Baselines}
Compared to using only the standard STA/LTA algorithm, Betelwise1 showed a significant increase in F1-score by [e.g., 35\%]. The hybrid system also outperformed the standalone conventional pipeline of Betelwise1, primarily due to the CNN's superior ability to distinguish subtle events from noise that mimics seismic signals.

\subsection{Ablation Study: Impact of Auxiliary Features}
To assess the contribution of the auxiliary statistical features (standard deviations before/after arrival) to the CNN, an ablation study was performed. A version of the CNN was trained using only the 1D waveform segment as input. The inclusion of auxiliary features resulted in an F1-score improvement of [e.g., 5-7\%], suggesting they provide valuable contextual information to the network.

\subsection{Computational Efficiency and Data Reduction}
The lightweight CNN model has a size of [e.g., X MB] and an average inference time of [e.g., Y milliseconds] per candidate event on [target hardware, e.g., a Raspberry Pi / specific CPU]. The overall Betelwise1 system, by selectively identifying high-confidence events, achieved a data reduction ratio of [e.g., 50:1 to 100:1] compared to transmitting continuous raw data, while successfully capturing over [e.g., 75\%] of cataloged events on the test datasets. This demonstrates its potential for optimizing telemetry bandwidth.

\section{Discussion}
The preliminary results from the Betelwise1 system are promising, suggesting that a hybrid approach combining conventional seismic processing with lightweight CNN refinement can be an effective strategy for efficient planetary seismic event detection.

The multi-stage design of Betelwise1 offers several advantages. The conventional pre-processing and initial picking stage acts as an efficient first-pass filter, significantly reducing the number of candidate events that need to be processed by the more computationally intensive CNN. The adaptive bandpass filtering and the characteristic function-based pick filtering were shown to be crucial in improving the quality of candidate picks. This tiered approach is particularly relevant for resource-constrained lander environments.

The use of 1D waveform segments augmented with auxiliary statistical features as input to the CNN is a key differentiator from approaches like that of \citet{civilini2021detecting}, which primarily used 2D spectrograms. Our findings, particularly the improvement seen with auxiliary features, suggest that providing direct temporal information alongside statistical context can be a viable and potentially more computationally efficient alternative for on-board processing, as it avoids the need for on-the-fly spectrogram generation for every candidate window. However, a direct, rigorous comparison of computational load and performance against an optimized spectrogram-based CNN on the same hardware would be necessary to definitively establish superiority for all scenarios.

The performance on lunar data indicates that Betelwise1 can generalize to planetary environments, even if primarily trained on terrestrial data or with limited planetary examples. The observed F1-score of [e.g., 0.81] on the Apollo PSE data demonstrates a practical capability for identifying moonquakes. Further testing and adaptation for Martian seismic data, considering its unique noise characteristics and event types (e.g., high-frequency (HF) and low-frequency (LF) marsquakes \citep{giardini2020seismicity}), will be critical.
